---
title: "Regular Expressions"
subtitle: "<br><br> STA35B: Statistical Data Science 2"
author: "Spencer Frei"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
    transition: faster
    css: cols.css
---


  
```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(unvotes)
library(knitr)
library(babynames)

hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
  lines <- options$output.lines
  if (is.null(lines)) {
    return(hook_output(x, options))  # pass to default hook
  }
  x <- unlist(strsplit(x, "\n"))
  more <- "..."
  if (length(lines)==1) {        # first n lines
    if (length(x) > lines) {
      # truncate the output, but add ....
      x <- c(head(x, lines), more)
    }
  } else {
    x <- c(more, x[lines], more)
  }
  # paste these lines together
  x <- paste(c(x, ""), collapse = "\n")
  hook_output(x, options)
})

knitr::opts_chunk$set(comment = NA) # makes it so the ## doesnt appear in output for chunks


```

---

* Data we will look at:
* `babynames` (use `install.packages("babynames")): year/sex/name/number/proportion vars
* `stringr::fruit`: 80 fruits
* `stringr::words`: 980 common English words
* `stringr::sentences`: 720 short sentences

We will use `str_view(string, pattern = NULL)` a lot.  `pattern` will parse regular expressions (regex)
```{r, output.lines=7}
str_view(fruit, "berry")
```



---




### Literal characters, metacharacters
* Letters and numbers which match exactly are **literal characters**
* Punctuation characters typically have special regex meanings (`.`, `+`, `*`, etc), and are called **metacharacters**
* The metacharacter `.` will match any character, so "a." matches any string which contains "a" followed by another character.
```{r}
str_view(c("a", "ab", "ae", "bd", "ea", "eab"), "a.")
```

* Or all fruits which have an "a", then 3 letters, then an "e":

```{r, output.lines=4}
str_view(fruit, "a...e")
```

---


### Quantifiers
* `?` makes a pattern optional (i.e. it matches 0 or 1 times)
* `+` lets a pattern repeat (i.e. it matches at least once)
* `*` lets a pattern be optional or repeat (i.e. it matches any number of times, including 0).

<div class="pull-left">
```{r}
# ab? matches an "a", optionally followed by "b".
str_view(c("a", "ab", "abb"), "ab?")

```

```{r}
# ab+ matches an "a", followed by at >=1 "b".
str_view(c("a", "ab", "abb"), "ab+")
```

</div>

<div class="pull-right">
```{r}
# ab* matches an "a", followed by any num of "b"s.
str_view(c("a", "ab", "abb"), "ab*")
```
</div>



---

### Character classes
* Defined by `[]`, let you match from a set of characters (similar idea to `%in%`)
  * `[abcd]` matches anything with "a", "b", "c", or "d"
  * Can negate/invert by using `^`: `[^abcd]` returns anything **except** "a", "b", "c", "d"
* e.g. any word containing "x" surrounded by vowels, or "y" surrounded by consonants
* **alternation** `|` picks between alternative patterns,  e.g. words containing "apple", "melon", or "nut"; repeated vowels

<div class="pull-left">
```{r}
str_view(words, "[aeiou]x[aeoiu]")
```
```{r}
str_view(words, "[^aeiou]y[^aeiou]")
```
</div>

<div class="pull-right">
```{r, output.lines=3}
str_view(fruit, "apple|melon|nut")
```
```{r}
str_view(fruit, "aa|ee|ii|oo|uu")
```
</div>


---

### `str_detect()`
* `str_detect(character_vector, pattern)` returns a logical vector, `TRUE` if pattern matches element of vector and `FALSE` otherwise.
```{r}
str_detect(c("a", "b", "c"), "[aeiou]")
```

<div class="pull-left">
* Since returns logical vectors, can be used with `filter()`, e.g. most popular names containing an "x":
```{r,output.lines=5}
babynames
```

</div>
<div class="pull-right">
```{r, output.lines=9}
babynames |> 
  filter(str_detect(name, "x")) |> 
  count(name, wt = n, sort = TRUE)
```

</div>

---

### `str_detect()`
* You can also use `str_detect()` in conjunction with `group_by()`, `summarize()` etc.
  * `sum()` will return number of strings which have pattern
  * `mean()` ill return proportion of strings which have pattern 
* E.g. proportion of names per year that have an "x"


```{r}
babynames %>% 
  group_by(year) %>%
  summarize(prop_x = mean(str_detect(name, "x"))) %>%
  arrange(by = desc(prop_x))
```

---

### Counting matches
* `str_count()` tells how many matches there are in a string
```{r}
x <- c("apple", "banana", "pear")
str_count(x, "p")
```
* Regex matches never overlap - always start after the end of previous match
```{r}
str_count("abababa", "aba")
```
```{r}
str_view("abababa", "aba")
```

---

### Counting vowels and constants in baby names
* Can use `str_count()` with mutate, i.e. computing number of vowels/consonants in baby names:

<div class="pull-left">
```{r}
babynames %>%
  count(name) %>%
  mutate(
    vowels = str_count(name, "[aeiou]"),
    consonants = str_count(name, "[^aeiou]")
  )
```
</div>

<div class="pull-right">
* Note that pattern matching is case sensitive, so "A" isn't counted.
* Ways around this:
  * Add the upper case vowels to the character class:     
    `str_count(name, "[aeiouAEIOU]")`
  * Use `str_to_lower()` to convert the names to lower case:
    `str_count(str_to_lower(name),   `
    `"[aeiou]")`
```{r, output.lines=4}
babynames %>% count(name) %>%  mutate(
    name = str_to_lower(name),
    vowels = str_count(name, "[aeiou]"),
    consonants = str_count(name, "[^aeiou]"))
```

</div>


---

### Replacing and removing values

* `str_replace()`: replaces first match 
* `str_replace_all()` replace all matches

```{r}
x <- c("apple", "pear", "banana")
(str_replace(x, "[aeiou]", "-"))
str_replace_all(x, "[aeiou]", "-")
```

* You can remove patterns if you set replacement with "", or using `str_remove()` / `str_remove_all()`

<div class="pull-left">
```{r}
str_remove(x, "[aeiou]")
```
</div>

<div class="pull-right">
```{r}
str_remove_all(x, "[aeiou]")
```
</div>

---

### Replacing characters and `?`, `*`, `+`
* The question mark `?` matches the preceding element zero OR ONE time, then iterates to rest of string.
* Plus sign `+` matches AT LEAST once
* The asterisk `*` matches the preceding element zero OR MORE times, then iterates to rest of string.

<div class="pull-left">
```{r}
x <- c("apple", "aardvark", "happy", "haaahaha")
( str_view(x, "a*") )
( str_remove(x, "a*") )
str_remove_all(x, "a*")
```
</div>

<div class="pull-right">
```{r}
( str_view(x, "a?") )
( str_remove(x, "a?") )
str_remove_all(x, "a?")
```


</div>

---

### Replacing characters and `?`, `*`, `+`
* Compare `?`, `+`, and `*`

<div class="pull-left">
```{r}
x <- c("apple", "aardvark", "happy", "haaahaha")
(str_view(x, "a?"))
(str_view(x, "a+"))
```

</div>

<div class="pull-right">
```{r}
(str_view(x, "a*"))
```
</div>

---

### Ranges of characters
* Suppose you have a vector of strings, and you want to do the following modifications:
  * If the string has a (lower/upper) letter between "a" and "u", replace it with an "x"
* Instead of spelling out manually what all of these letters are, you can use the character class operator `[]` together with `-`

<div class="pull-left">

An example with letters:
```{r}
x <- c("happy", "ab", "zap", "war")
( str_view(x, "[a-u]") )
str_replace_all(x, "[a-u]", "x")
```

</div>

<div class="pull-right">
An example with numbers: replace all numbers between 0 and 5 with x's
```{r}
x <- c("code9202", "apple2850", "0352")
(str_view(x, "[0-5]"))
str_replace_all(x, "[0-5]", "x")
```


---

### Ranges of characters and `?`, `*`

<div class="pull-left">
* Very useful to use ranges in conjunction with `?`, `*`, `+`
* E.g. let's find all words with at least three consecutive vowels
```{r}
str_view(words, "[aeiou][aeiou][aeiou]+")
```
* Useful for parsing strings which are partitioned by letters/numbers
</div>

<div class="pull-right">
```{r}
name_score <- c("Mary_92", "Pat_35", "Will_85")
( str_view(name_score, "[a-zA-Z]+")) 
str_view(name_score, "[0-9]+")
```
* E.g. replace all names with John, scores with 100
```{r}
name_score %>% str_replace("[a-zA-Z]+", "John") %>%
  str_replace("[0-9]+", "100")
```

<d/vi>
---

### Extracting variables
* `separate_wider_regex()`: go from long to wide using regex.

<div class="pull-left">
```{r}
df <- tribble(
  ~str,
  "<Sheryl>-F_34",
  "<Kisha>-F_45", 
  "<Pat>-X_33",
  "<Sharon>-F_38", 
  "<Penny>-F_58",
  "<Justin>-M_41", 
  "<Patricia>-F_84", 
)
```
* To extract data, construct sequence of regex that match each piece.
* If you want contents of that piece to appear in output, give it a name.
</div>

<div class="pull-right">
```{r}
df %>%  separate_wider_regex(
    str,
    patterns = c(
      "<", 
      name = "[A-Za-z]+", 
      ">-", 
      gender = ".",
      "_",
      age = "[0-9]+"))
```
</div>


---

### Escaping

<div class="pull-left">

* Since the characters ".", "?", "+", "*" have extra functions in regex, need to use escapes to help parse literal instances of these characters
* In regex, we require a `\` in front of characters to denote an escape
* But to create a string with an actual `\` in it, we need to use an escape, so need double `\\`:
```{r}
str_view(c("abc", "a.c", "bef"), "a\\.c")
```

```{r}
str_view(c("a*rdvark", "*pple", "m*n"), "\\*")
```

</div>

<div class="pull-right">
* Recall that to represent backslash in a string, need to escape:
```{r}
str_view("a\\b")
```
* To match for a backslash, need to create a string which has an escape in front of a backslash.
* The escape requires double backslash, and the string `\` also requires double backslash.

```{r}
str_view("a\\b", "\\\\")
```
* Backslashes are a nightmare. 
</div>

---


### Anchors

<div class="pull-left">
* By default: regex will match any part of a string.
* If you only want to match at beginning or end, you need to **anchor**:
  * `^` indicates "starts with" 
  * `$` indicates "ends with"

```{r}
str_view(fruit, "^a")

str_view(fruit, "a$")
```

</div>

<div class="pull-right">
* To force a regex to match only the full string, not subsets, anchor it with both `^` and `$`:
```{r}
str_view(fruit, "apple")
str_view(fruit, "^apple$")
```

* Example: replace every fruit name which starts with "a" with an "o"
```{r, output.lines=4}
str_replace(fruit, "^a", "o")
```

</div>



### Character sets

<div class="pull-left">

* We already saw how we can construct sets with `[]`: e.g. `[abc]` matches if any character is an "a", "b", or "c"
* We also saw how to use `-` to denote ranges, e.g. `[a-z]` lowercase letters, `[0-9]` numbers
* A few others:

  * `\d` matches any digit; `\D` matches anything that isn’t a digit.
  * `\s` matches any whitespace (e.g., space, tab, newline); `\S` matches anything that isn’t whitespace.
  * `\w` matches any “word” character, i.e. letters and numbers; `\W` matches any “non-word” character.

* Remember: to represent `\` in a string, need double backslash. 
</div>

<div class="pull-right">
```{r, eval=FALSE}
x <- "abcd ABCD 12345 -!@#%."
str_view(x, "\\d+")
#> [1] │ abcd ABCD <12345> -!@#%.
str_view(x, "\\D+")
#> [1] │ <abcd ABCD >12345< -!@#%.>
str_view(x, "\\s+")
#> [1] │ abcd< >ABCD< >12345< >-!@#%.
str_view(x, "\\S+")
#> [1] │ <abcd> <ABCD> <12345> <-!@#%.>
str_view(x, "\\w+")
#> [1] │ <abcd> <ABCD> <12345> -!@#%.
str_view(x, "\\W+")
#> [1] │ abcd< >ABCD< >12345< -!@#%.>
```

</div>



---


### Anchors: boundaries of words
* You can specify the beginning or end of the word using `\b` (i.e. "\\b" when typing a string)
  * This works by treating all letters and numbers as "word" characters, and everything else as "non-word" characters
```{r}
x <- c("summary(x)", "summarize(df)", "rowsum(x)", "sum(x)")
str_view(x, "sum")
str_view(x, "\\bsum\\b")
```

<div class="pull-left">


</div>

<div class="pull-right">
* 
</div>

---


### Quantifiers

<div class="pull-left">
* We already discussed `?` (0 or 1 match), `+` (1+ matches), `*` (0+ matches)
  * `colou?r`: matches American and British English
  * `\d+`: matches 1+ digits
  * `\s?`: matches 0+ whitespaces
* Can specify exact number of matches using `{}`: 


  * `{n}` matches exactly n times.
  *  `{n,}` matches at least n times.
  *  `{n,m}` matches between n and m times.

</div>

<div class="pull-right">
* Words with >= 3 consecutive vowels?
```{r, output.lines=5}
str_view(words, "[aeiou]{3,}")
```

* Words with between 4 and 6 consecutive consonants:
```{r}
str_view(words, "[^aeiou]{4,6}")
```

</div>

---

### Order of operations in regex

<div class="pull-left">

* Not immediately clear in which order R processes different operators.
  * `ab+`: is this "a" and then 1+ "b", or is it "ab" repeated 1+ times? (1st case)
  * `^a|b$`: match the string "a" or the string "b", OR:  string starting with "a" or string starting with "b" (2nd case)
* Generally: quantifiers (`?+*`) have high precedence, alternation `|` low.

</div>

<div class="pull-right">
* You can also introduce parenthesis to be more explicit about what you want, similar to normal math.

```{r}
str_view(words, "a(b+)") # same as `ab+`
```

```{r}
str_view(words, "(^a)|(b$)") # same as `^a|b$`
```
</div>

---

### Grouping and capturing with parenthesis

<div class="pull-left">
* With paranthesis, you can back-reference matches that appeared in parens, using `\1` for match in first parens, `\2` for match in second, etc.
* e.g. all fruits which have repeated **pair of letters**.
  * Pair of letters = "(..)"; back-ref: "\\1"
```{r}
str_view(fruit, "(..)\\1")
```
</div>

<div class="pull-right">

* Words that start and end with same pair of letters:
```{r}
# "starts with" a pair: ^(..)
# "ends with": need to end regex with \\1$
# to allow any chars between, put .* in middle
str_view(words, "^(..).*\\1$")
```

* Words that are repetitions of the same pair of letters:
```{r}
str_view(c("haha", "miumiu"), "^(..)+\\1$")
```

</div>

---

### More grouping and capturing

<div class="pull-left">
* Can also use back references in `str_replace()`, e.g. switching second and third words in sentences:
```{r, output.lines=7}
sentences |> 
  str_replace(
    "(\\w+) (\\w+) (\\w+)", 
    "\\1 \\3 \\2") |> 
  str_view()
```

</div>

<div class="pull-right">
* `(\\w+)`: matches with 1+ "word characters" (letters, numbers)
* Spacing between `(\\w+)` ensures we are looking for sequences of the form: word-chars, space, word-chars, space, word-chars
</div>

---

### Examples

<div class="pull-left">
* Words that start with "y":
```{r, output.lines=5}
str_view(words, "^y")
```

* Words that don't start with "y":
```{r,output.lines=5}
str_view(words, "^[^y]")
```

</div>

<div class="pull-right">

* Ends with a vowel-vowel-consonant triplet:
```{r, output.lines=5}
str_view(words, "[aeiou]{2}[^aeiou]$")
```

* Has 7 or more letters:
```{r, output.lines=5}
str_view(words, "[a-z]{7,}")
```

</div>


---

### Boolean operations

<div class="pull-left">
* We already saw how `^` inside `[]` negates the set, i.e. words with no vowels:
```{r,output.lines=5}
str_view(words, "^[^aeiou]+$")
```

* Another way: return vector of booleans indicating presence of vowels, then negate:

```{r,output.lines=5}
str_view(words[!str_detect(words, "[aeiou]")])
```

</div>

<div class="pull-right">
* This is useful since there's no "and" operator built into regex. 
* e.g., find all words that contain an "a" and a "b": trickier in standard regex,
```{r,output.lines=4}
str_view(words, "a.*b|b.*a")
```
* Easier with `str_detect()` and `&`:
```{r,output.lines=3}
str_view(words[str_detect(words, "a") 
               & str_detect(words, "b")])
```

</div>

---

### Boolean operations
* What if we wanted to find a word that contains "a", "e", "i", and "o"?
* If we tried to use standard regex, this would be very complex.
* Much easier using `str_detect()` and `&`:
```{r}
words[
  str_detect(words, "a") &
  str_detect(words, "e") &
  str_detect(words, "i") &
  str_detect(words, "o") 
]
```

<div class="pull-left">

</div>

<div class="pull-right">

</div>

---

### Creating patterns with code

<div class="pull-left">
* What if we want all sentences which mention a color?
  * Combine alternation with word boundaries `\b`:
```{r, output.lines=5}
str_view(sentences, "\\b(red|green|blue)\\b")
```
* But if we wanted to update this code to have more colors, would be pretty tedious to construct pattern by hand.
* We can build up larger regex's using functions we have seen before.
</div>

<div class="pull-right">
```{r,output.lines=5}
match_colors <- c("red","green", "blue")
for_regex <- str_c(
  "\\b(", 
  str_flatten(match_colors, "|"), 
  ")\\b")
str_view(sentences, for_regex)
```

* Then we can easily modify the list of colors by simply modifying `match_colors`. 
</div>

---

### Slide

<div class="pull-left">

</div>

<div class="pull-right">

</div>
