---
title: "Regular Expressions"
subtitle: "<br><br> STA35B: Statistical Data Science 2"
author: "Spencer Frei"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
    transition: faster
    css: cols.css
---


  
```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(unvotes)
library(knitr)
library(babynames)

hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
  lines <- options$output.lines
  if (is.null(lines)) {
    return(hook_output(x, options))  # pass to default hook
  }
  x <- unlist(strsplit(x, "\n"))
  more <- "..."
  if (length(lines)==1) {        # first n lines
    if (length(x) > lines) {
      # truncate the output, but add ....
      x <- c(head(x, lines), more)
    }
  } else {
    x <- c(more, x[lines], more)
  }
  # paste these lines together
  x <- paste(c(x, ""), collapse = "\n")
  hook_output(x, options)
})

knitr::opts_chunk$set(comment = NA) # makes it so the ## doesnt appear in output for chunks


```

---

* Data we will look at:
* `babynames` (use `install.packages("babynames")): year/sex/name/number/proportion vars
* `stringr::fruit`: 80 fruits
* `stringr::words`: 980 common English words
* `stringr::sentences`: 720 short sentences

We will use `str_view(string, pattern = NULL)` a lot.  `pattern` will parse regular expressions (regex)
```{r, output.lines=7}
str_view(fruit, "berry")
```



---




### Literal characters, metacharacters
* Letters and numbers which match exactly are **literal characters**
* Punctuation characters typically have special regex meanings (`.`, `+`, `*`, etc), and are called **metacharacters**
* The metacharacter `.` will match any character, so "a." matches any string which contains "a" followed by another character.
```{r}
str_view(c("a", "ab", "ae", "bd", "ea", "eab"), "a.")
```

* Or all fruits which have an "a", then 3 letters, then an "e":

```{r, output.lines=4}
str_view(fruit, "a...e")
```

---


### Quantifiers
* `?` makes a pattern optional (i.e. it matches 0 or 1 times)
* `+` lets a pattern repeat (i.e. it matches at least once)
* `*` lets a pattern be optional or repeat (i.e. it matches any number of times, including 0).

<div class="pull-left">
```{r}
# ab? matches an "a", optionally followed by "b".
str_view(c("a", "ab", "abb"), "ab?")

```

```{r}
# ab+ matches an "a", followed by at >=1 "b".
str_view(c("a", "ab", "abb"), "ab+")
```

</div>

<div class="pull-right">
```{r}
# ab* matches an "a", followed by any num of "b"s.
str_view(c("a", "ab", "abb"), "ab*")
```
</div>



---

### Character classes
* Defined by `[]`, let you match from a set of characters (similar idea to `%in%`)
  * `[abcd]` matches anything with "a", "b", "c", or "d"
  * Can negate/invert by using `^`: `[^abcd]` returns anything **except** "a", "b", "c", "d"
* e.g. any word containing "x" surrounded by vowels, or "y" surrounded by consonants
* **alternation** `|` picks between alternative patterns,  e.g. words containing "apple", "melon", or "nut"; repeated vowels

<div class="pull-left">
```{r}
str_view(words, "[aeiou]x[aeoiu]")
```
```{r}
str_view(words, "[^aeiou]y[^aeiou]")
```
</div>

<div class="pull-right">
```{r, output.lines=3}
str_view(fruit, "apple|melon|nut")
```
```{r}
str_view(fruit, "aa|ee|ii|oo|uu")
```
</div>


---

### `str_detect()`
* `str_detect(character_vector, pattern)` returns a logical vector, `TRUE` if pattern matches element of vector and `FALSE` otherwise.
```{r}
str_detect(c("a", "b", "c"), "[aeiou]")
```

<div class="pull-left">
* Since returns logical vectors, can be used with `filter()`, e.g. most popular names containing an "x":
```{r,output.lines=5}
babynames
```

</div>
<div class="pull-right">
```{r, output.lines=9}
babynames |> 
  filter(str_detect(name, "x")) |> 
  count(name, wt = n, sort = TRUE)
```

</div>

---

### `str_detect()`
* You can also use `str_detect()` in conjunction with `group_by()`, `summarize()` etc.
  * `sum()` will return number of strings which have pattern
  * `mean()` ill return proportion of strings which have pattern 
* E.g. proportion of names per year that have an "x"


```{r}
babynames %>% 
  group_by(year) %>%
  summarize(prop_x = mean(str_detect(name, "x"))) %>%
  arrange(by = desc(prop_x))
```

---

### Counting matches
* `str_count()` tells how many matches there are in a string
```{r}
x <- c("apple", "banana", "pear")
str_count(x, "p")
```
* Regex matches never overlap - always start after the end of previous match
```{r}
str_count("abababa", "aba")
```
```{r}
str_view("abababa", "aba")
```

---

### Counting vowels and constants in baby names
* Can use `str_count()` with mutate, i.e. computing number of vowels/consonants in baby names:

<div class="pull-left">
```{r}
babynames %>%
  count(name) %>%
  mutate(
    vowels = str_count(name, "[aeiou]"),
    consonants = str_count(name, "[^aeiou]")
  )
```
</div>

<div class="pull-right">
* Note that pattern matching is case sensitive, so "A" isn't counted.
* Ways around this:
  * Add the upper case vowels to the character class:     
    `str_count(name, "[aeiouAEIOU]")`
  * Use `str_to_lower()` to convert the names to lower case:
    `str_count(str_to_lower(name),   `
    `"[aeiou]")`
```{r, output.lines=4}
babynames %>% count(name) %>%  mutate(
    name = str_to_lower(name),
    vowels = str_count(name, "[aeiou]"),
    consonants = str_count(name, "[^aeiou]"))
```

</div>


---

### Replacing and removing values

* `str_replace()`: replaces first match 
* `str_replace_all()` replace all matches

```{r}
x <- c("apple", "pear", "banana")
(str_replace(x, "[aeiou]", "-"))
str_replace_all(x, "[aeiou]", "-")
```

* You can remove patterns if you set replacement with "", or using `str_remove()` / `str_remove_all()`

<div class="pull-left">
```{r}
str_remove(x, "[aeiou]")
```
</div>

<div class="pull-right">
```{r}
str_remove_all(x, "[aeiou]")
```
</div>

