---
title: "Regular Expressions"
subtitle: "<br><br> STA35B: Statistical Data Science 2"
author: "Spencer Frei"
output:
  xaringan::moon_reader:
    # lib_dir: libs
    nature:
      ratio: "16:9"
      # highlightLines: true
      # highlightStyle: solarized-light
      countIncrementalSlides: false
---
  
```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(unvotes)
library(knitr)

hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
  lines <- options$output.lines
  if (is.null(lines)) {
    return(hook_output(x, options))  # pass to default hook
  }
  x <- unlist(strsplit(x, "\n"))
  more <- "..."
  if (length(lines)==1) {        # first n lines
    if (length(x) > lines) {
      # truncate the output, but add ....
      x <- c(head(x, lines), more)
    }
  } else {
    x <- c(more, x[lines], more)
  }
  # paste these lines together
  x <- paste(c(x, ""), collapse = "\n")
  hook_output(x, options)
})

knitr::opts_chunk$set(comment = NA) # makes it so the ## doesnt appear in output for chunks
```

Data we will look at:
* `babynames` (use `install.packages("babynames")): year/sex/name/number/proportion vars
* `stringr::fruit`: 80 fruits
* `stringr::words`: 980 common English words
* `stringr::sentences`: 720 short sentences

We will use `str_view(string, pattern = NULL)` a lot.  `pattern` will parse regular expressions (regex)
```{r, eval = FALSE}
str_view(fruit, "berry")
#>  [6] │ bil<berry>
#>  [7] │ black<berry>
#> [10] │ blue<berry>
#> [11] │ boysen<berry>
#> [19] │ cloud<berry>
#> [21] │ cran<berry>
#> ... and 8 more
```



---




### Literal characters, metacharacters
* Letters and numbers which match exactly are **literal characters**
* Punctuation characters typically have special regex meanings (`.`, `+`, `*`, etc), and are called **metacharacters**
* The metacharacter `.` will match any character, so "a." matches any string which contains "a" followed by another character.
```{r, eval=FALSE}
str_view(c("a", "ab", "ae", "bd", "ea", "eab"), "a.")
#> [2] │ <ab>
#> [3] │ <ae>
#> [6] │ e<ab>
```
* Or all fruits which have an "a", then 3 letters, then an "e":
```{r, eval=FALSE}
str_view(fruit, "a...e")
#>  [1] │ <apple>
#>  [7] │ bl<ackbe>rry
#> [48] │ mand<arine>
#> [51] │ nect<arine>
#> [62] │ pine<apple>
#> [64] │ pomegr<anate>
#> ... and 2 more
```