---
title: "Visualizations IV / Factors"
subtitle: "<br><br> STA35B: Statistical Data Science 2"
author: "Spencer Frei"
output:
  xaringan::moon_reader:
    # lib_dir: libs
    nature:
      ratio: "16:9"
      # highlightLines: true
      # highlightStyle: solarized-light
      countIncrementalSlides: false
---
  


  
```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(unvotes)
library(knitr)

hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
  lines <- options$output.lines
  if (is.null(lines)) {
    return(hook_output(x, options))  # pass to default hook
  }
  x <- unlist(strsplit(x, "\n"))
  more <- "..."
  if (length(lines)==1) {        # first n lines
    if (length(x) > lines) {
      # truncate the output, but add ....
      x <- c(head(x, lines), more)
    }
  } else {
    x <- c(more, x[lines], more)
  }
  # paste these lines together
  x <- paste(c(x, ""), collapse = "\n")
  hook_output(x, options)
})

knitr::opts_chunk$set(comment = NA) # makes it so the ## doesnt appear in output for chunks

source("../_common.R")
```

### Factors
.pull-left[ 
* Factors are used for categorical variables that have a fixed and known set of values
* Useful when we want to display character vectors in non-alphabetical order
* E.g. let's suppose we have a variable that records month:
```{r}
x1 <- c("Dec", "Apr", "Jan", "Mar")
```
Two issues:
* Only twelve possible months, so if we typed "Jam" instead of "Jan", would be hard to catch the error.
* Sorting by characters is not meaningful:
```{r}
sort(x1)
```

]

.pull-right[
* We can fix these by treating them as a factor
* Factors require a list of valid **levels**:
```{r}
month_levels <- c(
  "Jan", "Feb", "Mar", "Apr", "May", "Jun", 
  "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"
)
```
* With levels, you can create factors:
```{r}
y1 <- factor(x1, levels = month_levels)
y1
sort(y1)
```

]


---

### Factors
.pull-left[
* If you create a factor from a vector, every element of the vector must be a level of a factor, otherwise you get `NA`:
```{r}
x2 <- c("Dec", "Apr", "Jam", "Mar")
(y2 <- factor(x2, levels = month_levels))
```
* If you want to throw an error for non-level vector info, you can use ``forcats::fct()`
```{r, eval = FALSE}
y2 <- fct(x2, levels = month_levels)
#> Error in `fct()`:
#> ! All values of `x` must appear in `levels` or `na`
#> ℹ Missing level: "Jam"
```

]

.pull-right[
* If you omit levels, `factor()` makes factors in alphabetical order:
```{r}
factor(x1)
```
* `forcats::fct()` orders by first appearance in the vector
```{r}
fct(x1)
```
* You can extract levels from a factor using `levels()`:
```{r}
levels(y2)
```

]

---

### Factor: General Social Survey
* We'll look at survey data from the General Social Survey (`forcats::gss_cat`, loaded in tidyverse)
```{r}
gss_cat
```
* When factors are in a tibble, levels do not directly show, but can see them using `count`:

.pull-left[
```{r, eval=FALSE}
gss_cat %>%
  count(race)
```
]

.pull-right[
```{r, echo=FALSE}
gss_cat %>%
  count(race)
```
]


---

### GSS Data: exploratory data analysis
.pull-left[ 
* Let's look at the distribution of `rincome` (reported income)
```{r}
#| fig-width: 6
ggplot(gss_cat, aes(x = rincome)) + geom_bar()
```
* A little hard to see what's happening on the x-axis!
]

--

.pull-right[


```{r}
#| fig-width: 6
ggplot(gss_cat, aes(y = rincome)) + geom_bar()
```

]

---

### GSS Data: exploratory data analysis
.pull-left[ 
* Does income vary by religion much?
```{r}
#| fig-width: 6
ggplot(gss_cat, aes(x = rincome)) + geom_bar()
```
* A little hard to see what's happening on the x-axis!
]

--

.pull-right[


```{r}
#| fig-width: 6
ggplot(gss_cat, aes(y = rincome)) + geom_bar()
```

]

---

### Exploratory data analysis
.pull-left[ 
* What are the most common religions in the survey?  What about party ID?  

```{r}
gss_cat %>% count(relig) %>% arrange(n) %>% print(n=Inf)
```

]

.pull-right[

```{r}
gss_cat %>% count(partyid) %>% arrange(n) %>% print(n=Inf)
```

]
---

.pull-left[ 
* Let's look at different TV watching habits by marital status.  Maybe we can first try doing a `geom_freqpoly()` with color given by religion.
```{r, message=FALSE, warning=FALSE}
ggplot(gss_cat,
       aes(x = tvhours, y = after_stat(density), color = marital)) + 
  geom_freqpoly()
```
]

--

.pull-right[
* If we try the same thing for religion, it is trickier to visualize:

* Let's look at TV hours watched per religion.  Maybe we can first try doing a `geom_freqpoly()` with color given by religion.

```{r, message=FALSE, warning=FALSE}
ggplot(gss_cat,
       aes(x = tvhours, y = after_stat(density), color = relig)) + 
  geom_freqpoly()
```

* We need to reduce the amount of data we're trying to show.
]

---
.pull-left[ 
* To reduce amount of information, we should summarize the data, e.g. by taking the average per religion.

```{r}
#| fig-width: 5
relig_summary <- gss_cat |>
  group_by(relig) |>
  summarize(
    tvhours = mean(tvhours, na.rm = TRUE),
    n = n()
  )

ggplot(relig_summary, aes(x = tvhours, y = relig)) + 
  geom_point()
```
* Hard to read because there's no clear pattern in the plot.

]

.pull-right[
* To re-order, we can improve it by reordering using `fact_reorder( f=, x=, fun=)`
* `f`: factor whose levels to modify
* `x`: numeric vector to give the new order of levels
* `fun`: function to use if multiple values of `x` for given value of `f` (default: `median`)
```{r}
#| fig-width: 5
ggplot(relig_summary,
       aes(x = tvhours, y = fct_reorder(relig, tvhours))) +
  geom_point()
```

]

---

### Modifying factor levels 

.pull-left[
* Often will want to change values of levels so that plots look better
```{r}
gss_cat %>% count(partyid)
```
* Not ideal for plotting - inconsistent (acronyms vs full spelling etc)
* Key function: `fct_recode()`

]

.pull-left[
```{r}
gss_cat %>%
  mutate(
    partyid = fct_recode(partyid,
      "Republican, strong"    = "Strong republican",
      "Republican, weak"      = "Not str republican",
      "Independent, near rep" = "Ind,near rep",
      "Independent, near dem" = "Ind,near dem",
      "Democrat, weak"        = "Not str democrat",
      "Democrat, strong"      = "Strong democrat")) %>%
  count(partyid)
```
* Levels not explicitly mentioned will be kept as is. 
]



---

### Modifying factor levels

.pull-left[
* To combine groups, assign multiple old levels to same new level
```{r, eval = FALSE}
gss_cat |>
  mutate(
    partyid = fct_recode(partyid,
      "Republican, strong"    = "Strong republican",
      "Republican, weak"      = "Not str republican",
      "Independent, near rep" = "Ind,near rep",
      "Independent, near dem" = "Ind,near dem",
      "Democrat, weak"        = "Not str democrat",
      "Democrat, strong"      = "Strong democrat",
      "Other"                 = "No answer",
      "Other"                 = "Don't know",
      "Other"                 = "Other party"
    )
  )
```

]

.pull-right[
* If you are collapsing many levels, can use `fct_collapse`:
```{r}
gss_cat |>
  mutate(
    partyid = fct_collapse(partyid,
      "other" = c("No answer", "Don't know", "Other party"),
      "rep" = c("Strong republican", "Not str republican"),
      "ind" = c("Ind,near rep", "Independent", "Ind,near dem"),
      "dem" = c("Not str democrat", "Strong democrat")
    )
  ) |>
  count(partyid)
```
]


---

### Examples
* Let's now try and put some of all of the ideas we've seen so far together
* Look at the `weather` tibble in nycflights13:
```{r}
library(nycflights13)
weather
```
* Let's see if we can come up with a way of visualizing the temperature trends across every year.
* We want to visualize what are the minimum temperature, maximum temperature, and average temperature per month, at each location

---

### Data prep
* We need to compute the min/max/average temp per month in every year in every location
```{r, warning=FALSE,message=FALSE}
weather2 <- weather %>% 
  group_by(year, month, origin) %>%
  summarize(min_temp = min(temp, na.rm=TRUE), max_temp = max(temp, na.rm=TRUE),
            avg_temp = mean(temp, na.rm=TRUE)) %>%
  group_by(month, origin) %>%
  summarize(min_temp = mean(min_temp, na.rm=TRUE),
            max_temp = mean(max_temp, na.rm=TRUE),
            avg_temp = mean(avg_temp, na.rm=TRUE), n = n())
```
* This tibble only has one year in it, so we don't actually need the code following the second group by.
* However, if there are > 1 years in tibble, this code will compute the average min/max/mean temp per month across years.

---


.pull-left[ 

* Let's try plotting the monthly temps at Newark (EWR).
```{r}
#| fig-width: 5
weather2 %>% filter(origin == 'EWR') %>%
  ggplot(aes(x = month)) + 
  geom_line(aes(y = min_temp), color = 'blue') +
  geom_line(aes(y = max_temp), color = 'red') + 
  geom_line(aes(y = avg_temp), color = 'black')
```
* This is OK, but not ideal.  No legend appears by default, also have to manually add three geom_line's.
* Better idea is to clean the tibble so that we have groups of data: for this we need data to be *long*
]

.pull-right[
```{r}
(weather3 <- weather2 %>%
  pivot_longer(cols = c(min_temp, max_temp, avg_temp),
               values_to = "temperature",
               names_to = "measurement")
)
```


]

---
### Visualizing
.pull-left[
* Now we can do color by temperature type:
```{r}
ggplot(weather3 %>% filter(origin == 'EWR'),
       aes(x = month, y = temperature, color = measurement)) +
  geom_line()
```
* Now we'd like better legend; blue for min, black for avg, and red for max; fix ticks 
]

.pull-left[
```{r}
ggplot(weather3 %>% filter(origin == 'EWR'),
       aes(x = month, y = temperature, color = measurement)) +
  geom_line(linewidth = 2) + 
  scale_color_manual(
    values = c(min_temp = "blue", avg_temp = "black", max_temp = "red"),
    labels = c("min_temp" = "minimum", "avg_temp" = "average", "max_temp" = "maximum")) + 
  scale_x_continuous(breaks = 1:12, minor_breaks = 1:12) + 
  labs(title = "Temperature at EWR per month")
```
]

---
### Visualizing
.pull-left[ 
* Now let's do the same type of plot but with each airport's min/avg/max temp per month plotted side-by-side.
```{r}
#| fig-width: 8
#| fig-height: 2
#| eval: false
ggplot(weather3,
       aes(x = month, y = temperature, color = measurement)) +
  geom_line(linewidth = 2) + 
  scale_color_manual(
    values = c(min_temp = "blue", avg_temp = "black", max_temp = "red"),
    labels = c("min_temp" = "minimum", "avg_temp" = "average", "max_temp" = "maximum")) + 
  scale_x_continuous(breaks = 1:12, minor_breaks = 1:12) + 
  facet_grid(. ~ origin)
```
]

.pull-right[
```{r}
#| fig-width: 8
#| echo: false
ggplot(weather3,
       aes(x = month, y = temperature, color = measurement)) +
  geom_line(linewidth = 2) + 
  scale_color_manual(
    values = c(min_temp = "blue", avg_temp = "black", max_temp = "red"),
    labels = c("min_temp" = "minimum", "avg_temp" = "average", "max_temp" = "maximum")) + 
  scale_x_continuous(breaks = 1:12, minor_breaks = 1:12) + 
  facet_grid(. ~ origin)
```

]

* What happened at JFK in May?  Why is it so different from EWR and LGA?

---

.pull-left[ 

```{r}
weather %>%
  filter(month == 5, temp < 25)
```
* Appears real!

]

.pull-right[

* How does it compare to EWR and LGA? 
```{r}
weather %>% filter(month == 5, day == 8) %>%
  filter(between(hour, 21, 23)) %>% print(n = Inf)
```

* Appears to be only a single instance of this, so definitely an outlier, can't be certain for why it happens.

]



---

### Midterm studying tips
* Review labs, homeworks (and homework solutions - HW 2 sol up tonight), and practice midterm
* Make sure you understand all of the core functions: `min`, `max`, `mean`, `pmin`, `pmax`, `group_by()`, `summarize()`, `pivot_wider`, `pivot_longer`, regex, joins, etc.   Understand how NA's work, what typical default behavior for NA's is, etc. 
* Test-taking strategy: Go through exam and solve all of the easy questions last.  If it takes more than 2 minutes, skip and return later.
* Work through examples on the margins / side / back of the exam to make sure you're understanding things correctly.