---
title: 'Lab 4'
author: ''
output:
  pdf_document
---


```{r set-up, echo = FALSE}
knitr::opts_chunk$set(warning = FALSE, 
                      echo = TRUE, 
                      message = FALSE)
```


```{r load-packages, warning=FALSE, message=FALSE}
library(tidyverse)
library(palmerpenguins)
library(ggthemes) 
```


 

Refer to the dataset 'penguins',

(1) Drop all rows with missing values (1 pt).

(2) Use `geom_point()`, draw a scatter plot between x = 'body_mass_g' and y ='bill_length_mm', colored by 'island', shaped by 'sex' (2 pts).

(3) Use `geom_smooth()`, draw a line to describe the relationship between x = 'body_mass_g' and y ='bill_length_mm', remove the confidence interval around smooth, set color as black and size as 0.5 (2 pts).

(4) Use `facet_grid()`,  split the plot into subplots based on 'species' horizontally (1 pt).
 
(5) Use `scale_color_manual()`, change the colors of Biscoe, Dream and Torgersen  as blue, black, and red respectively (1 pt).

(6) Use `scale_shape_manual()`, change the shapes of female and male as #1 and #3 respectively, and change 'female' and 'male' to 'Female' and 'Male' in the legends (2 pts).

(7) Use `scale_x_continuous()`, change the x-axis scales as 3,4,5,6 (1 pt).

(8) Use `labs()`, set x-axis label as "Body Mass (kg)", y-axis label as "Bill Length (mm)", color label as "Island", shape label as "Sex", subtitle as 'Species', and title as "Scatterplot of Body Mass v.s. Bill Length" (2 pts).

(9) Use `theme()`, place the legend at the left-hand side of the plot, center both th plot title and subtitle (2 pts).

(10) Use `theme()`, bold the legend titles and the plot title (1 pt).

```{r}
# code here 
```

