---
title: 'Lab 6'
author: ''
output:
  pdf_document
---


```{r set-up, echo = FALSE}
knitr::opts_chunk$set(warning = FALSE, 
                      echo = TRUE, 
                      message = FALSE)
```


```{r load-packages, warning=FALSE, message=FALSE}
library(tidyverse)
library(MASS)
library(ggthemes) 
```


 
### Question 1

Refer to the dataset 'iris',

(1) Use `str`, find the data types of variables in iris.

(2) Use `levels`, find the levels of 'Species' variable in iris.

(3) Fit a linear model 'm.full' of 'Sepal.Length' on all remaining variables and get the summary. 

(4) Use 'update' to do backward elimination.  Is there a smaller model with better adjusted R squared?  What is the final model after using backward elimination? 


### Question 2

(1) Use `rnorm` to generate 100 samples from $$N(0,1)$$, and name it as 'sple'.

(2) Calculate the mean of 'sple', and name it as 'm'.

(3) Draw 100 bootstrap samples and calculate their means. Name the means as 'ms'.

(4) Use 'ggplot' to draw a histogram of 'ms'.

(5) Using these 100 bootstrap samples, provide an estimate for the 95% bootstrap confidence interval for the population mean.
