---
title: "Dates and times"
subtitle: "<br><br> STA35B: Statistical Data Science 2"
author: "Spencer Frei"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
    transition: faster
    css: cols.css
---


  
```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(unvotes)
library(knitr)
library(babynames)

hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
  lines <- options$output.lines
  if (is.null(lines)) {
    return(hook_output(x, options))  # pass to default hook
  }
  x <- unlist(strsplit(x, "\n"))
  more <- "..."
  if (length(lines)==1) {        # first n lines
    if (length(x) > lines) {
      # truncate the output, but add ....
      x <- c(head(x, lines), more)
    }
  } else {
    x <- c(more, x[lines], more)
  }
  # paste these lines together
  x <- paste(c(x, ""), collapse = "\n")
  hook_output(x, options)
})

knitr::opts_chunk$set(comment = NA) # makes it so the ## doesnt appear in output for chunks


```

---

### Dates and times: complications

A number of things contribute to dates and times being more complex than one might think.

* Not all years have 365 days
  * The actual rule that determines leap years:
```{}
A year is a leap year if it’s divisible by 4, 
unless it’s also divisible by 100, 
except if it’s also divisible by 400. 
In other words, in every set of 400 years, there’s 97 leap years.
```
* Not every day in every location has 24 hours a day
  * Daylight savings time implies one day has 23, another has 24
* Time zones are difficult!
* We will be using **lubridate** package (part of *latest* tidyverse), and nycflights13.
```{r}
library(tidyverse)
library(nycflights13)
```

---

### Creating dates and times

<div class="pull-left">

Three types of date/time data:

  * A **date**. Tibbles print this as `<date>`.
  * A **time** within a day. Tibbles print this as `<time>`.
  * A **date-time** is a date plus a time: it uniquely identifies an instant in time (typically to the nearest second). Tibbles print this as `<dttm>`. Base R calls these POSIXct, but doesn’t exactly trip off the tongue.

* We are going to focus on dates and date-times as R doesn’t have a native class for storing times. If you need one, you can use the **hms** package.

</div>

<div class="pull-right">

To get the current date or date-time you can use today() or now():
```{r}
today()
```
```{r}
now()
(class(now()))
```
</div>


---

### Dates and times from strings

<div class="pull-left">
* Number of functions which create dates from strings: three letter combos of "y", "m", "d"
```{r}
ymd("2017-01-31")
```
```{r}
mdy("January 31st, 2017")
```
```{r}
mdy("January 31, 2017")
```
```{r}
dmy("31-Jan-2017")
```


</div>

<div class="pull-right">
* To create date-times, you can add an underscore and then one or more of "h", "m", "s".
```{r}
(ymd_hms("2017-01-31 20:11:59"))
mdy_hm("01/31/2017 08:01")
```
* Times are assumed to be UTC time zone; can change by using `tz=`
```{r}
mdy_hm("01/31/2017 08:01", tz = "PST")
```

</div>

---

### Creating date-times from dplyr parts
Remember how flights stored some of the date information:
```{r, output.lines=5}
(flights_select <- flights %>% select(
  year, month, day, hour, minute))
```
* To create date/time from this, can use `make_date()` or `make_datetime()`:
```{r, output.lines=5}
flights_select %>%
  mutate(departure = make_datetime(year, month, day, hour, minute))
```

<div class="pull-left">

</div>

<div class="pull-right">

</div>

---

### Creating date-times from dplyr parts

* We'll now do a similar computation for the four time columns in flights
* We'll do so using a **function** - we haven't seen this yet, but we will see it in a couple weeks 
```{r, output.lines=4}
make_datetime_100 <- function(year, month, day, time) {
  make_datetime(year, month, day, time %/% 100, time %% 100)
} # recall: time is an integer, use modular arithmetic to convert 

flights_dt <- flights |> 
  filter(!is.na(dep_time), !is.na(arr_time)) |> 
  mutate(
    dep_time = make_datetime_100(year, month, day, dep_time),
    arr_time = make_datetime_100(year, month, day, arr_time),
    sched_dep_time = make_datetime_100(year, month, day, sched_dep_time),
    sched_arr_time = make_datetime_100(year, month, day, sched_arr_time)
  ) |> 
  select(origin, dest, ends_with("delay"), ends_with("time"))

flights_dt
```



<div class="pull-left">

</div>

<div class="pull-right">

</div>

---

### Updated flights df with times for arrivals/departures
* We'll now use this update df
```{r}
flights_dt %>%
  filter(dep_time < ymd(20130102))
```


<div class="pull-left">

</div>

<div class="pull-right">

</div>

---

### Date-time components

<div class="pull-left">
* Accessor functions which are helpful for date-time types: 
  * `year()`, `month()`,  `hour()`, `minute()`, and `second()`
  * `mday()` (day of the month), `yday()` (day of the year), `wday()` (day of the week)
  
```{r}
datetime <- ymd_hms("2026-07-08 12:34:56")
year(datetime)
month(datetime)
mday(datetime)
```

</div>

<div class="pull-right">
```{r}
yday(datetime)
wday(datetime) # 2026-07-08 is Weds. (Sun.=1)
```

* `month()` and `wday()` can have `label=TRUE`, returns abbreviated name of month/day
* Set `abbr=FALSE` to get full name
```{r}
datetime <- ymd_hms("2026-07-08 12:34:56")
month(datetime, label=TRUE) 
wday(datetime, label=TRUE, abbr = FALSE)
```


</div>

---

### Date-time components
With this you can do things like calculate the minute with the highest departure delays:
```{r}
flights_dt %>%
  mutate(minute = minute(dep_time)) %>%
  group_by(minute) %>%
  summarize(avg_delay = mean(dep_delay, na.rm=TRUE)
  ) %>% arrange(by = desc(avg_delay))
```
<div class="pull-left">

</div>

<div class="pull-right">


</div>

---

### Rounding
* There are analogues of the standard rounding functions for dates
  * `floor_date()`, `ceiling_date()`
  * `round_date()`
* They take vector of dates to adjust, name of unit (week, day, etc)

<div class="pull-left">
```{r, output.lines=10}
flights_dt %>%
  mutate(year = floor_date(dep_time, "year")) %>%
  select(dep_time, year)
```

</div>

<div class="pull-right">

</div>

---

### Examples
* Let's compute the average delay time of flights which depart at times in two groups:
  * departure time is between minutes 20-30 and 50-60 vs. the other times 
<div class="pull-left">
```{r}
flights_dt %>%
  mutate(dep_minute = minute(dep_time),
         mins_2030 = dep_minute >= 20 & dep_minute <= 30,
         mins_5060 = dep_minute >= 50 & dep_minute <= 59,
         mins_2030_or_5060 = mins_2030 | mins_5060) %>%
  group_by(mins_2030_or_5060) %>%
  summarize(avg_dep_delay = mean(dep_delay, na.rm=TRUE),
            n = n())
  
```

</div>

<div class="pull-right">

</div>

---

### Time spans and date-time arithmetic

<div class="pull-left">
Three important classes representing time spans:

* **Durations**: exact time, measured in seconds
* **Periods**: human units, like weeks/months
* **Intervals**: represent starting/end point


</div>

<div class="pull-right">
* When subtracting two dates, we get a "difftime" object:
```{r}
# How old is Hadley?
h_age <- today() - ymd("1979-10-14")
h_age
```
* Exact unit of difftime can vary from seconds, minutes, hours, days, or weeks.
* `as.duration()` always uses seconds:
```{r}
as.duration(h_age)
```


</div>

---

### Constructing durations

<div class="pull-left">
* Durations always record time span in seconds
* Useful constructors: `d{units}`, `{units}` is `seconds`, `days`, etc
  * No way to convert month to duration since it's not well-defined
```{r}
dseconds(15)
dminutes(10)
dhours(c(12, 24))
```

</div>

<div class="pull-right">
* Can add and multiply durations:
```{r}
2 * dyears(1)
dyears(1) + dweeks(12) + dhours(5)
```
* Can subtract durations to and from days
```{r}
(tomorrow <- today() + ddays(1)) # returns Date
(last_year <- today() - dyears(1)) # returns date-time
```

</div>

---

### Duration computations and weird results
* Durations represent an exact number of seconds
* So if you are looking at time zones, things can be odd
```{r}
one_am <- ymd_hms("2026-03-08 01:00:00", tz = "America/New_York")
one_am
one_am + ddays(1)
```
* If we add a full day of seconds, would need to account for hour time change from EST to EDT.
* lubridate provides **periods** to address this
<div class="pull-left">

</div>

<div class="pull-right">

</div>

---

### Periods

<div class="pull-left">
* Periods are time spans but don't have fixed length in seconds
  * Work like "human" times, i.e. days/months
```{r}
one_am
one_am + days(1)
one_am + ddays(1)
```

</div>

<div class="pull-right">
* Similar to durations, useful constructors and behavior under `+`/`*`:
```{r}
hours(c(12, 24))
days(7)
months(1:3)
10 * (months(6) + days(2))
```

</div>

---

### Durations vs. periods


<div class="pull-left">
* Adding periods can be a bit more in line with expectations
```{r, eval=FALSE}
# A leap year
ymd("2024-01-01") + dyears(1)
#> [1] "2024-12-31 06:00:00 UTC"
ymd("2024-01-01") + years(1)
#> [1] "2025-01-01"

# Daylight saving time
one_am + ddays(1)
#> [1] "2026-03-09 02:00:00 EDT"
one_am + days(1)
#> [1] "2026-03-09 01:00:00 EDT"
```

</div>

<div class="pull-right">

</div>

---

### Fixing a bug in `flights_dt`
* We used same date information for arrival and departure times, but the flights really arrived on the following day.
```{r, output.lines=5}
flights_dt %>% filter(arr_time < dep_time) %>% select(origin, dest, arr_time, dep_time)
```

```{r}
flights_dt %>% mutate(
  overnight = arr_time < dep_time, # returns T/F
  arr_time = arr_time + days(overnight),
  sched_arr_time = sched_arr_time + days(overnight)
) %>% filter(arr_time < dep_time)
```

<div class="pull-left">

</div>

<div class="pull-right">

</div>

---

### Intervals

<div class="pull-left">
* `dyears(1) / ddays(365)` does not return 1, since `dyears()` is defined as the number of seconds per average year: 365.25 days.
* `years(1) / days(1)` does not return 365, since in leap years this isn't true.
* **Intervals** allow for defining specific intervals of time, using pair of starting/end date times.
* Format: `start %--% end`:
</div>

<div class="pull-right">
```{r}
(y2023 <- ymd("2023-01-01") %--% ymd("2024-01-01"))
(y2024 <- ymd("2024-01-01") %--% ymd("2025-01-01"))
```
* Can divide by `days()` to find out how many days/year:
```{r}
y2023 / days(1)
y2024 / days(1)
```

</div>

---

### Example: extracting dates and computing durations
* Suppose we are given a tibble whose first three rows are as follows:
```{r}
 df <- tribble(
    ~name, ~entry,
    "Mary", "First arrival: January 2, 2005; Second arrival: January 6, 2023",
    "Will", "First time: January 5, 1997; Second time: January 8, 2015",
    "Jose", "First visit: January 4, 1990; Second visit: January 9, 2008",
  )
```
* Task: return a tibble which says how many days elapsed between the first and second visit, ordered by the most number of days
* Complex task!  Need to:
  * Parse the "entry" column to extract the dates
  * Turn them into proper date-times
  * Compute the number of days between visits (leap years happen at different times!)
  * Order by number of days

* Let's start by parsing "entry" and creating two columns for different dates
<div class="pull-left">

</div>

<div class="pull-right">
</div>

---

* `entry` elements look like: First arrival: January 2, 2005; Second arrival: January 9, 2023"
```{r}
( df2 <- df %>% mutate(
    date1 = mdy(str_replace(entry, "(.*): (.*); (.*): (.*)", "\\2")),
    date2 = mdy(str_replace(entry, "(.*): (.*); (.*): (.*)", "\\4"))
  ) )
```
* Now we want to compute number of days between: `ddays` of the difference!
```{r}
df2 %>% mutate(days_elapsed = days(date2 - date1) )
```

<div class="pull-left">
</div>

<div class="pull-right">

</div>

---

### Examples

* Just need to order by maximum number of days now 
```{r}
df %>% mutate(
    date1 = mdy(str_replace(entry, "(.*): (.*); (.*): (.*)", "\\2")),
    date2 = mdy(str_replace(entry, "(.*): (.*); (.*): (.*)", "\\4")),
    days_elapsed = days(date2-date1)
  ) %>% arrange(by = desc(days_elapsed))
```

* We could also have done `separate_wider_regex()` to parse the string initially. How would we do that?
<div class="pull-left">

</div>

<div class="pull-right">

</div>

---

### Examples


<div class="pull-left">
```{r}
df2 <- df %>% separate_wider_regex(
  entry,
  patterns = c(
    ".*",
    ": ",
    date1 = ".*",
    ";",
    ".*",
    ": ",
    date2 = ".*"
  ) ) %>%
  mutate(date1 = mdy(date1), date2 = mdy(date2))
```
* Original df:
```{r}
df
```

</div>

<div class="pull-right">
```{r}
df2
```

</div>

---

### Example - string formatting errors for dates
* Let's suppose we are given a string with a bunch of supposed dates
```{r}
ds <- c("2022-01-08", "202-01-09", "2022/01/10")
```
* And suppose we want to return a new datetime vector satisfying the following:
  * if the date is in the correct format, it is the original datetime
  * otherwise, return the datetime of january 1, year 1. 
* We need to flag which elements of `ds` have the format of either YYYY-MM-DD or YYYY/MM/DD, i.e. return a vector with booleans saying whether that element has the correct format.
* We'll use regular expressions
```{r}
pattern_regex <- "^\\d{4}[-/]\\d{2}[-/]\\d{2}$"
str_detect(ds, pattern_regex)
```

<div class="pull-left">

</div>

<div class="pull-right">

</div>

---

### Example - string formatting errors for dates

```{r}
ds
str_detect(ds, "^\\d{4}[-/]\\d{2}[-/]\\d{2}$")
```
* Now we can use a `case_when` to make it year 1 if not correct format:
```{r}
df <- tibble(entered = ds)
df %>% mutate(new_date = case_when(
  str_detect(entered, "^\\d{4}[-/]\\d{2}[-/]\\d{2}$") ~ ymd(entered, quiet=TRUE),
  TRUE ~ ymd("0001-01-01") )
  )
```


<div class="pull-left">

</div>

<div class="pull-right">

</div>

---

### Slide

<div class="pull-left">

</div>

<div class="pull-right">

</div>

---

### Slide

<div class="pull-left">

</div>

<div class="pull-right">

</div>

---

### Slide

<div class="pull-left">

</div>

<div class="pull-right">

</div>

---

### Slide

<div class="pull-left">

</div>

<div class="pull-right">

</div>

---

### Slide

<div class="pull-left">

</div>

<div class="pull-right">

</div>

---

### Slide

<div class="pull-left">

</div>

<div class="pull-right">

</div>

---
