---
title: "Dates and times"
subtitle: "<br><br> STA35B: Statistical Data Science 2"
author: "Spencer Frei"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
    transition: faster
    css: cols.css
---


  
```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(unvotes)
library(knitr)
library(babynames)

hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
  lines <- options$output.lines
  if (is.null(lines)) {
    return(hook_output(x, options))  # pass to default hook
  }
  x <- unlist(strsplit(x, "\n"))
  more <- "..."
  if (length(lines)==1) {        # first n lines
    if (length(x) > lines) {
      # truncate the output, but add ....
      x <- c(head(x, lines), more)
    }
  } else {
    x <- c(more, x[lines], more)
  }
  # paste these lines together
  x <- paste(c(x, ""), collapse = "\n")
  hook_output(x, options)
})

knitr::opts_chunk$set(comment = NA) # makes it so the ## doesnt appear in output for chunks


```

---

### Dates and times: complications

A number of things contribute to dates and times being more complex than one might think.

* Not all years have 365 days
  * The actual rule that determines leap years:
```{}
A year is a leap year if it’s divisible by 4, 
unless it’s also divisible by 100, 
except if it’s also divisible by 400. 
In other words, in every set of 400 years, there’s 97 leap years.
```
* Not every day in every location has 24 hours a day
  * Daylight savings time implies one day has 23, another has 24
* Time zones are difficult!
* We will be using **lubridate** package (part of *latest* tidyverse), and nycflights13.
```{r}
library(tidyverse)
library(nycflights13)
```

---

### Creating dates and times

<div class="pull-left">

Three types of date/time data:

  * A **date**. Tibbles print this as `<date>`.
  * A **time** within a day. Tibbles print this as `<time>`.
  * A **date-time** is a date plus a time: it uniquely identifies an instant in time (typically to the nearest second). Tibbles print this as `<dttm>`. Base R calls these POSIXct, but doesn’t exactly trip off the tongue.

* We are going to focus on dates and date-times as R doesn’t have a native class for storing times. If you need one, you can use the **hms** package.

</div>

<div class="pull-right">

To get the current date or date-time you can use today() or now():
```{r}
today()
```
```{r}
now()
(class(now()))
```
</div>


---

### Dates and times from strings

<div class="pull-left">
* Number of functions which create dates from strings: three letter combos of "y", "m", "d"
```{r}
ymd("2017-01-31")
```
```{r}
mdy("January 31st, 2017")
```
```{r}
mdy("January 31, 2017")
```
```{r}
dmy("31-Jan-2017")
```


</div>

<div class="pull-right">
* To create date-times, you can add an underscore and then one or more of "h", "m", "s".
```{r}
(ymd_hms("2017-01-31 20:11:59"))
mdy_hm("01/31/2017 08:01")
```
* Times are assumed to be UTC time zone; can change by using `tz=`
```{r}
mdy_hm("01/31/2017 08:01", tz = "PST")
```

</div>

---

### Creating date-times from dplyr parts
Remember how flights stored some of the date information:
```{r, output.lines=5}
(flights_select <- flights %>% select(
  year, month, day, hour, minute))
```
* To create date/time from this, can use `make_date()` or `make_datetime()`:
```{r, output.lines=5}
flights_select %>%
  mutate(departure = make_datetime(year, month, day, hour, minute))
```

<div class="pull-left">

</div>

<div class="pull-right">

</div>

---

### Slide

<div class="pull-left">

</div>

<div class="pull-right">

</div>

---

### Slide

<div class="pull-left">

</div>

<div class="pull-right">

</div>

---

### Slide

<div class="pull-left">

</div>

<div class="pull-right">

</div>

---

### Slide

<div class="pull-left">

</div>

<div class="pull-right">

</div>

---

### Slide

<div class="pull-left">

</div>

<div class="pull-right">

</div>

---

### Slide

<div class="pull-left">

</div>

<div class="pull-right">

</div>

---

### Slide

<div class="pull-left">

</div>

<div class="pull-right">

</div>

---

### Slide

<div class="pull-left">

</div>

<div class="pull-right">

</div>

---

### Slide

<div class="pull-left">

</div>

<div class="pull-right">

</div>

---

### Slide

<div class="pull-left">

</div>

<div class="pull-right">

</div>

---

### Slide

<div class="pull-left">

</div>

<div class="pull-right">

</div>

---

### Slide

<div class="pull-left">

</div>

<div class="pull-right">

</div>

---

### Slide

<div class="pull-left">

</div>

<div class="pull-right">

</div>

---

### Slide

<div class="pull-left">

</div>

<div class="pull-right">

</div>

---

### Slide

<div class="pull-left">

</div>

<div class="pull-right">

</div>

---

### Slide

<div class="pull-left">

</div>

<div class="pull-right">

</div>

---

### Slide

<div class="pull-left">

</div>

<div class="pull-right">

</div>

---

### Slide

<div class="pull-left">

</div>

<div class="pull-right">

</div>

---

### Slide

<div class="pull-left">

</div>

<div class="pull-right">

</div>

---

### Slide

<div class="pull-left">

</div>

<div class="pull-right">

</div>

---

### Slide

<div class="pull-left">

</div>

<div class="pull-right">

</div>

---

### Slide

<div class="pull-left">

</div>

<div class="pull-right">

</div>

---

### Slide

<div class="pull-left">

</div>

<div class="pull-right">

</div>

---

### Slide

<div class="pull-left">

</div>

<div class="pull-right">

</div>

---

### Slide

<div class="pull-left">

</div>

<div class="pull-right">

</div>

---
