---
title: "Visualizations III"
subtitle: "<br><br> STA35B: Statistical Data Science 2"
author: "Spencer Frei"
output:
  xaringan::moon_reader:
    # lib_dir: libs
    nature:
      ratio: "16:9"
      # highlightLines: true
      # highlightStyle: solarized-light
      countIncrementalSlides: false
---
  


  
```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(unvotes)
library(knitr)

hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
  lines <- options$output.lines
  if (is.null(lines)) {
    return(hook_output(x, options))  # pass to default hook
  }
  x <- unlist(strsplit(x, "\n"))
  more <- "..."
  if (length(lines)==1) {        # first n lines
    if (length(x) > lines) {
      # truncate the output, but add ....
      x <- c(head(x, lines), more)
    }
  } else {
    x <- c(more, x[lines], more)
  }
  # paste these lines together
  x <- paste(c(x, ""), collapse = "\n")
  hook_output(x, options)
})

knitr::opts_chunk$set(comment = NA) # makes it so the ## doesnt appear in output for chunks

source("../_common.R")
```


### Further inspecting diamond prices
* Recall that it appeared that diamond prices seemed inversely correlated with quality
```{r}
#| fig-alt: 
#| fig-width: 4

ggplot(diamonds, aes(x = cut, y = price)) +
  geom_boxplot()
```
* What might be responsible for this? Recall variables: x,y,z refer to length/width/depth of diamond in mm
```{r}
colnames(diamonds)
```

* Let's look at **size** vs price.


---

### Size vs. price in diamonds
* We can try doing 3 side-by-side plots by just doing 3 ggplots:
```{r}
#| layout-ncol: 3
#| fig-width: 3.5
ggplot(diamonds, aes(x = x, y = price)) +
  geom_point()
ggplot(diamonds, aes(x = y, y = price)) +
  geom_point()
ggplot(diamonds, aes(x = z, y = price)) +
  geom_point()
```
* Some things we notice:
  * Clearly some outliers - "zero" length / width / depth diamonds apparently?
  * But still some positive correlation between x, y, z and price
  * Hard to see with so many dots
* Let's use `alpha=` to add some transparency, make it easier on the eyes

---

### Size vs. price in diamonds
* We can try doing 3 side-by-side plots by just doing 3 ggplots:
```{r}
#| layout-ncol: 3
#| fig-width: 3.5
ggplot(diamonds, aes(x = x, y = price)) +
  geom_point(alpha = 0.02)
ggplot(diamonds, aes(x = y, y = price)) +
  geom_point(alpha = 0.02)
ggplot(diamonds, aes(x = z, y = price)) +
  geom_point(alpha = 0.02)
```
* Some things we notice:
  * Clearly some outliers - "zero" length / width / depth diamonds apparently?
  * But still some positive correlation between x, y, z and price
  * Hard to see with so many dots
* Let's now clean up the tibble so we are only looking at the middle 95% of values for x, y, z

---

### Size vs. price in diamonds
* Let's clean up the tibble so we are only looking at the middle 95% of values for x, y, z
```{r}
#| layout-ncol: 3
#| fig-width: 4
diamonds_middle <- diamonds %>%
  filter(between(x, quantile(diamonds$x, 0.025), quantile(diamonds$x, 0.975)),
         between(y, quantile(diamonds$y, 0.025), quantile(diamonds$y, 0.975)),
         between(z, quantile(diamonds$z, 0.025), quantile(diamonds$z, 0.975)))
ggplot(diamonds_middle, aes(x = x, y = price)) +
  geom_point(alpha = 0.02)
ggplot(diamonds_middle, aes(x = y, y = price)) +
  geom_point(alpha = 0.02)
ggplot(diamonds_middle, aes(x = z, y = price)) +
  geom_point(alpha = 0.02)
```

---
### Size vs. price in diamonds
* Appears x, y, z (length/width/depth) are highly correlated with price
* Recall that volume of a cube is length * width * depth - so let's look at the relationship between estimated volume (x\*y\*z) and price
```{r}
#| layout-ncol: 1
#| fig-width: 5
diamonds_middle %>%
  mutate(est_volume = x*y*z) %>%
  ggplot(aes(x = est_volume, y = price)) +
  geom_point(alpha = 0.01)
```
* Does appear to be a strong correlation.
* Now let's think again as to why it was that we previously found that higher quality diamonds were cheaper.

---
### Size vs. price in diamonds
* Hypothesis: higher quality diamonds are smaller, lower quality diamonds are larger.
* To investigate, we could use boxplot: we want to visualize variation in size for each category of quality
.pull-left[ 
```{r}
diamonds_middle %>%
  mutate(est_volume = x*y*z) %>%
  ggplot(aes(x = cut, y = est_volume)) +
  geom_boxplot()
```

]

---

### Size vs price in diamonds
* Let's try to visualize the different cuts on the dot plot:
```{r}
#| layout-ncol: 2
#| fig-width: 5
diamonds_middle %>%
  mutate(est_volume = x*y*z) %>%
  ggplot(aes(x = est_volume, y = price, color = cut)) +
  geom_point(alpha = 0.5)
diamonds_middle %>%
  mutate(est_volume = x*y*z) %>%
  ggplot(aes(x = est_volume, y = price, color = cut)) +
  geom_smooth()
```


---

### Size vs. price in diamonds
* We were looking at the estimated volume, but could also look at the *weight*: carat.  
* Let's see how if weight varies by class.

.pull-left[ 
```{r}
#| layout-ncol: 1
#| fig-width: 4
diamonds %>%
  ggplot(aes(x = cut, y = carat)) +
  geom_boxplot()
```
]

.pull-right[
* Again appears to be a lot of outliers; let's look at middle 95%.
```{r}
#| layout-ncol: 1
#| fig-width: 4
diamonds_middle <- diamonds %>%
  filter(between(carat, quantile(carat, 0.025), quantile(carat, 0.975)))
diamonds_middle %>%
  ggplot(aes(x = cut, y = carat)) +
  geom_boxplot()
```

]
---
### Size vs price in diamonds 
.pull-left[ 
* And now let's visualize relationship between carat size and price by cut:
```{r, echo=FALSE, message=FALSE}
#| layout-ncol: 1
#| fig-width: 4.5
diamonds_middle %>%
  ggplot(aes(x = carat, y = price, color = cut)) +
  geom_point(alpha = 0.5, show.legend=FALSE)
diamonds_middle %>%
  mutate(est_volume = x*y*z) %>%
  ggplot(aes(x = carat, y = price, color = cut)) +
  geom_smooth()
```
]
.pull-right[
* Compare this to what happens if we didn't ignore outliers!
```{r, echo=FALSE, message=FALSE}
#| layout-ncol: 1
#| fig-width: 4.5
diamonds %>%
  ggplot(aes(x = carat, y = price, color = cut)) +
  geom_point(alpha = 0.5, show.legend=FALSE)
diamonds %>%
  mutate(est_volume = x*y*z) %>%
  ggplot(aes(x = carat, y = price, color = cut)) +
  geom_smooth()
```

]

---

### Communicating ideas
* Now we'll look at how to best display data once we've figured out some meaningful relationships between things
* Packages we'll be using: `tidyverse`, `scales`, `ggrepel`, `patchwork`
```{r, echo=FALSE,message=FALSE}
library(tidyverse)
library(scales)
library(ggrepel)
library(patchwork)
```
.pull-left[
* Good labels are crucial to good figures:
```{r, eval = FALSE}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  labs(
    x = "Engine displacement (L)",
    y = "Highway fuel economy (mpg)",
    color = "Car type",
    title = "Fuel efficiency generally decreases with engine size",
    subtitle = "Two seaters (sports cars) are an exception because of their light weight",
    caption = "Data from fueleconomy.gov"
  )
```

]

.pull-right[
```{r, echo=FALSE, message=FALSE}
#| fig-width: 5.5
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  labs(
    x = "Engine displacement (L)",
    y = "Highway fuel economy (mpg)",
    color = "Car type",
    title = "Fuel efficiency generally decreases with engine size",
    subtitle = "Two seaters (sports cars) are an exception because of their light weight",
    caption = "Data from fueleconomy.gov"
  )
```

]


---

### Annotations
.pull-left[ 
* Often helpful to label individual observations or gruops of observations
* Tool to use: `geom_text()`, similar to `geom_point` but has additional aesthetic `label`.
* Sources of labels:
  - Tibble itself has labels
  - Custom, user-added labels.
```{r}
label_info <- mpg |>
  group_by(drv) |>
  arrange(desc(displ)) |>
  slice_head(n = 1) |>
  mutate(
    drive_type = case_when(
      drv == "f" ~ "front-wheel drive",
      drv == "r" ~ "rear-wheel drive",
      drv == "4" ~ "4-wheel drive"
    )
  ) |>
  select(displ, hwy, drv, drive_type)
```

  ]
  
.pull-right[
```{r, eval=FALSE}
label_info
#> # A tibble: 3 × 4
#> # Groups:   drv [3]
#>   displ   hwy drv   drive_type       
#>   <dbl> <int> <chr> <chr>            
#> 1   6.5    17 4     4-wheel drive    
#> 2   5.3    25 f     front-wheel drive
#> 3   7      24 r     rear-wheel drive
```
* Can use this to directly label groups and replace legend
]
  
---

### Annotations
.pull-left[
```{r}
label_info
```
* Want to use `label_info` as the data.
* x and y aesthetics the same as in the original `mpg` tibble
* But then want to set `hjust` and `vjust` 
]

.pull-right[
```{r}
#| fig-width: 4.5
ggplot(mpg, aes(x = displ, y = hwy, color = drv)) +
  geom_point(alpha = 0.3) +
  geom_smooth(se = FALSE) +
  geom_text(
    data = label_info, 
    aes(x = displ, y = hwy, label = drive_type),
    fontface = "bold", size = 5, hjust = "right", vjust = "bottom"
  ) +
  theme(legend.position = "none")
```

]


---

### Annotations
.pull-left[
```{r, echo=FALSE, message = FALSE}
#| fig-width: 4.5
ggplot(mpg, aes(x = displ, y = hwy, color = drv)) +
  geom_point(alpha = 0.3) +
  geom_smooth(se = FALSE) +
  geom_text(
    data = label_info, 
    aes(x = displ, y = hwy, label = drive_type),
    fontface = "bold", size = 5, hjust = "right", vjust = "bottom"
  ) +
  theme(legend.position = "none")
```
* Two things to address: (1) What are `hjust` and `vjust` doing, why is text appearing where it does? (2) 
* Let's look at `label_info`
```{r, eval = FALSE}
label_info
#>   displ   hwy drv   drive_type       
#>   <dbl> <int> <chr> <chr>            
#> 1   6.5    17 4     4-wheel drive    
#> 2   5.3    25 f     front-wheel drive
#> 3   7      24 r     rear-wheel drive
```

]

.pull-right[
* We can use ``ggrepel::geom_label_repel()` to make text repel other text labels.
```{r}
#| fig-width: 4
ggplot(mpg, aes(x = displ, y = hwy, color = drv)) +
  geom_point(alpha = 0.3) +
  geom_smooth(se = FALSE) +
  geom_label_repel(
    data = label_info, 
    aes(x = displ, y = hwy, label = drive_type),
    fontface = "bold", size = 5, nudge_y = 2
  ) +
  theme(legend.position = "none")
```

]

---

### Annotations
.pull-left[ 
* Same idea can be used to highlight points on plot, e.g. outliers
```{r, eval=FALSE}
potential_outliers <- mpg |>
  filter(hwy > 40 | (hwy > 20 & displ > 5))
  
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  geom_text_repel(data = potential_outliers, aes(label = model)) +
  geom_point(data = potential_outliers, color = "red") +
  geom_point(
    data = potential_outliers,
    color = "red", size = 3, shape = "circle open"
  )
```

]
.pull-right[
```{r, echo=FALSE}
potential_outliers <- mpg |>
  filter(hwy > 40 | (hwy > 20 & displ > 5))
  
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  geom_text_repel(data = potential_outliers, aes(label = model)) +
  geom_point(data = potential_outliers, color = "red") +
  geom_point(
    data = potential_outliers,
    color = "red", size = 3, shape = "circle open"
  )
```

]

---
### Annotations
* Other annotations which you might encounter / could be useful:
  * `geom_hline()`, `geom_vline()`: create horizontal / vertical lines
  * `geom_rect()`: rectangle around points of interest, require specifying `xmin`, `xmax`, `ymin`, `ymax`
  * `geom_segment()`: draw an arrow from one point to another
* Function which allows for doing any of these: `annotate()`
* Let's say we want to add some text to our plot, but split lines every 30 characters using `str_wrap()`
```{r}
trend_text <- "Larger engine sizes tend to have lower fuel economy." |>
  str_wrap(width = 30)
trend_text
```

---

### Annotations

.pull-left[

* Let's add a text label in red using `annotate(geom = 'label')`
* ... and then add a line segment giving general trend

```{r, eval=FALSE}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  annotate(
    geom = "label", x = 3.5, y = 38,
    label = trend_text,
    hjust = "left", color = "red"
  ) +
  annotate(
    geom = "segment",
    x = 3, y = 35, xend = 5, yend = 25, color = "red",
    arrow = arrow(type = "closed")
  )
```

]

.pull-right[
```{r, echo=FALSE}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  annotate(
    geom = "label", x = 3.5, y = 38,
    label = trend_text,
    hjust = "left", color = "red"
  ) +
  annotate(
    geom = "segment",
    x = 3, y = 35, xend = 5, yend = 25, color = "red",
    arrow = arrow(type = "closed")
  )
```

]

---

### Scales
* By default, ggplot2 automatically adds scales for you, e.g. this code
```{r, eval=FALSE}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(color = class))
```
uses the default behavior for scaling `geom_point()`, which is using continuous scales for x and y and discrete color scales:

```{r, eval = FALSE}
#| fig-width: 4.5
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(color = class)) +
  scale_x_continuous() +
  scale_y_continuous() +
  scale_color_discrete()
```

---

### Axis ticks
* Axes and legends are called **guides** in R.  Axes used for x,y aesthetics, legends for everything else
* Ticks on axes and keys on legend affected by args `breaks` and `labels`.

.pull-left[
```{r}
#| fig-width: 4
ggplot(mpg, aes(x = displ, y = hwy, color = drv)) +
  geom_point() +
  scale_y_continuous(breaks = seq(15, 40, by = 5)) 
```

]

.pull-right[
```{r}
#| fig-width: 4
ggplot(mpg, aes(x = displ, y = hwy, color = drv)) +
  geom_point() +
  scale_x_continuous(labels = NULL) +
  scale_y_continuous(labels = NULL) +
  scale_color_discrete(labels = c("4" = "4-wheel", "f" = "front", "r" = "rear"))
```

]

---

### Axis ticks
* Specifying `labels` and `breaks` gives a lot of flexibility on how to plot

```{r}
#| fig-width: 4.5
#| layout-ncol: 2
# Left
ggplot(diamonds, aes(x = price, y = cut)) +
  geom_boxplot(alpha = 0.05) +
  scale_x_continuous(labels = label_dollar())

# Right
ggplot(diamonds, aes(x = price, y = cut)) +
  geom_boxplot(alpha = 0.05) +
  scale_x_continuous(
    labels = label_dollar(scale = 1/1000, suffix = "K"), 
    breaks = seq(1000, 19000, by = 6000) )
```

---

### Legends
.pull-left[ 
* To control location of legend, use `theme()` (themes control non data parts of a plot)
```{r}
#| layout-ncol: 2
#| fig-width: 4
#| eval: false

base <- ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(color = class))

base + theme(legend.position = "right") # the default
base + theme(legend.position = "left")
base + 
  theme(legend.position = "top") +
  guides(color = guide_legend(nrow = 3))
base + 
  theme(legend.position = "bottom") +
  guides(color = guide_legend(nrow = 3))
```
]

.pull-right[
```{r}
#| layout-ncol: 2
#| fig-width: 3
#| echo: false

base <- ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(color = class))

base + theme(legend.position = "right") # the default
base + theme(legend.position = "left")
base + 
  theme(legend.position = "top") +
  guides(color = guide_legend(nrow = 3))
base + 
  theme(legend.position = "bottom") +
  guides(color = guide_legend(nrow = 3))
```

]


