---
title: "Functions 1"
subtitle: "<br><br> STA35B: Statistical Data Science 2"
author: "Spencer Frei"
output:
  xaringan::moon_reader:
    # lib_dir: libs
    nature:
      ratio: "16:9"
      # highlightLines: true
      # highlightStyle: solarized-light
      countIncrementalSlides: false
---
  


  
```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(unvotes)
library(knitr)

hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
  lines <- options$output.lines
  if (is.null(lines)) {
    return(hook_output(x, options))  # pass to default hook
  }
  x <- unlist(strsplit(x, "\n"))
  more <- "..."
  if (length(lines)==1) {        # first n lines
    if (length(x) > lines) {
      # truncate the output, but add ....
      x <- c(head(x, lines), more)
    }
  } else {
    x <- c(more, x[lines], more)
  }
  # paste these lines together
  x <- paste(c(x, ""), collapse = "\n")
  hook_output(x, options)
})

knitr::opts_chunk$set(comment = NA) # makes it so the ## doesnt appear in output for chunks

source("../_common.R")
```


### Iteration
* We've already seen a few ways to do computations repeatedly in a clean / easy way:
  - `facet_grid()` and `facet_wrap()`, drawing plots for each group
  - `group_by()` + `summarize()` for summarys tatistics for each group
  - creating custom functions
  
* We'll see what other ways R allows for us to flexibly do calculations and save time

---

###  Modifying multiple columns

.pull-left[
* Consider a simple tibble (`runif(n)`: n independent standard normals), and that we want to compute median of every column:
```{r}
df <- tibble(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)
df |> summarize(
  n = n(),
  a = median(a),
  b = median(b),
  c = median(c),
  d = median(d),
)
```

]

.pull-right[
* Should never be copying+pasting more than twice (what if we had 500 columns!)
* Helpful function: `across()`:
```{r}
df %>%
  summarize(
    n = n(),
    across(a:d, median)
  )
```
* In coming slides, we'll see `across` works and how to modify this behavior.
* Three especially important arguments to `across()`:
  - `.cols`: which columns to iterate over
  - `.fns`: what to do (function) ofr each column
  - `.names`: name output of each column
]

---

### `across()`: Selecting columns with `.cols`

.pull-left[
* For `.cols`, we can use same things we used for `select()`:
```{r}
df %>% summarize(across(-a, median))
df %>% summarize(across(c(a,c), median))
```
* Two additional arguments which are helpful: `everything()` and `where()`. 
```{r}
df %>% summarize(across(everything(), median))
```

]
.pull-right[
* `everything()` computes summaries for every non-grouping variable
* `where()` allows for selecting columns based on type, e.g. `where(is.numeric)` for numbers, `where(is.character)` for strings, `where(is.logical)` for logicals, etc.
```{r}
df <- tibble(
  grp = sample(2, 10, replace = TRUE), # either 1 or 2
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10))
df |> 
  group_by(grp) |> 
  summarize(across(everything(), median))
```

]

---

### `across()`: calling a single function
* `.fns` says how we want data to be transformed
* We are passing the *function* to `across()`, we are not calling the function itself. 
  - Never add the `()` after the function when you pass to across, otherwise you get an error.
```{r, eval=FALSE}
df |> 
  group_by(grp) |> 
  summarize(across(everything(), median()))
#> Error in `summarize()`:
#> ℹ In argument: `across(everything(), median())`.
#> Caused by error in `median.default()`:
#> ! argument "x" is missing, with no default
```
* Same reason why calling `median()` in console will result in an error, since it has no input.

---
### `across()`: calling multiple functions 
* What happens if you want to do multiple 
